---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
### Jonathan Hill

## Loading necessary R packages
```{r}
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

I'll name my dataset "activity" and convert the date variable from a factor to a date.

```{r}
activity <- read.csv("activity.csv",header=TRUE)
activity$date <- as.Date(activity$date)
```

Then process the data so that it is summarized by the day and calculate the mean and median.

```{r}
activityDay <- activity %>%
        group_by(date) %>%
        summarize (stepsDay = sum(steps)
                   )
meanstepsperday <- mean(activityDay$stepsDay,na.rm=TRUE)
medianstepsperday <- median(activityDay$stepsDay,na.rm=TRUE)
```

## What is mean total number of steps taken per day?

The mean total number of steps taken per day is `r meanstepsperday` and the median is `r medianstepsperday`.

### Histogram of steps per day

```{r}
ggplot(data=activityDay,mapping = aes(date,stepsDay))+
        geom_bar(stat="identity", fill ="dark blue")+
        labs(y="Steps per Day", x="Date", title="Histogram of Steps per Day")
```

## What is the average daily activity pattern?

First I will reshape the data and summarize it by the day. 

```{r}
activityInt <- activity %>%
        group_by(interval) %>%
        summarize (stepsInt = sum(steps, na.rm=TRUE)/61
        )
```

Then graph it here:
```{r}
ggplot(data =activityInt,mapping=aes(interval,stepsInt))+
        geom_line()+
        labs(y="Average steps per interval",x="5-Minute Interval",title="Time Series Graph")
```

Next, I calculate the 5-minute interval in which the maximum number of steps are taken.
```{r}
maxInt <- activityInt$interval[which.max(x = activityInt$stepsInt)]
```

The 5-minute interval in which the maximum average steps are taken is `r maxInt`.

## Imputing missing values
```{r}
missingval <- sum(is.na(activity$steps))
```

The dataset has `r missingval` missing values. Therfore, I will merge it with the dataset from the average daily activity pattern above.  Then I will use an ifelse statement to replace the missing values with the average number of steps in a given 5-minute interval.
```{r}
mergedactivity <- merge(activityInt,activity,by="interval")
mergedactivity$steps <- ifelse(is.na(mergedactivity$steps),
                         mergedactivity$stepsInt,
                         mergedactivity$steps)
```

Finally, in order to bring the data back to its original state, I drop the average calculation variable.
```{r}
mergedactivity$stepsInt <- NULL
```

Here is the new histogram.
```{r}
cleanhist <- mergedactivity %>%
        group_by(date) %>%
        summarize (stepsDay = sum(steps)
        )

ggplot(data=cleanhist,mapping = aes(date,stepsDay))+
        geom_bar(stat="identity", fill ="dark blue")+
        labs(y="Steps per Day", x="Date", title="Histogram of Steps per Day")
```
```{r}
cleanmean <- mean(cleanhist$stepsDay)
cleanmedian <- median(cleanhist$stepsDay)
```

The new mean is `r cleanmean` and the new median is `r cleanmedian`.  By imputing the missing values, the mean decreases by `r meanstepsperday - cleanmean` and the median decreases by `r medianstepsperday - cleanmedian`.

## Are there differences in activity patterns between weekdays and weekends?

Creating a new factor variable to indicate weekend or weekday:
```{r}
mergedactivity$dow <- weekdays(mergedactivity$date)
mergedactivity$weekend <- ifelse(mergedactivity$dow == "Saturday","weekend",ifelse(mergedactivity$dow == "Sunday","weekend","weekday"))
mergedactivity$weekend <- as.factor(mergedactivity$weekend)
```

Manipulating the data
```{r}
cleanInt <- mergedactivity %>%
        group_by(interval,weekend) %>%
        summarize (stepsInt = mean(steps, na.rm=TRUE)
        )
```

Creating the faceted line graphs
```{r}
ggplot(data =cleanInt,mapping=aes(interval,stepsInt))+
        geom_line()+
        facet_grid(weekend~.)+
        labs(y="Average steps per interval",x="5-Minute Interval",title="Faceted Time Series Graph")
```
